---
title: "Genomics Technologies"
# linktitle: "1. Introduction to RStudio and Images"
date: "2026-01-12"
class_date: "2026-01-12"
draft: false
output:
  blogdown::html_page:
    toc: false
menu:
  class:
    parent: Class sessions
    weight: 1
type: docs
weight: 1
# pdf: 
# thumb: 
editor_options: 
  chunk_output_type: console
---

## Welcome slides

<a href="../../slides/01_Genomic_technologies/01_Intro.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/01_Genomic_technologies/01_Intro.html#1')
```

## Genomics technologies

<a href="../../slides/01_Genomic_technologies/02a_Genome.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/01_Genomic_technologies/02a_Genome.html#1')
```

a href="../../slides/01_Genomic_technologies/02b_Early_Sequencing.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/01_Genomic_technologies/02b_Early_Sequencing.html#1')
```

a href="../../slides/01_Genomic_technologies/02c_Human_Genome.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/01_Genomic_technologies/02c_Human_Genome.html#1')
```

a href="../../slides/01_Genomic_technologies/02d_Modern_Sequencing.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/01_Genomic_technologies/02d_Modern_Sequencing.html#1')
```

a href="../../slides/01_Genomic_technologies/02e_Third_Generation.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/01_Genomic_technologies/02e_Third_Generation.html#1')
```

a href="../../slides/01_Genomic_technologies/02f_Applications.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/01_Genomic_technologies/02f_Applications.html#1')
```
