---
title: "Scientific Computing"
# linktitle: "1. Introduction to RStudio and Images"
date: "2026-01-21"
class_date: "2026-01-21"
draft: false
output:
  blogdown::html_page:
    toc: false
menu:
  class:
    parent: Class sessions
    weight: 2
type: docs
weight: 2
# pdf: 
# thumb: 
editor_options: 
  chunk_output_type: console
---

## Unix

<a href="../../slides/02a_Unix/01_Unix_xaringan.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/02a_Unix/01_Unix_xaringan.html#1')
```

## Text

<a href="../../slides/02a_Unix/02_Text_xaringan.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/02a_Unix/02_Text_xaringan.html#1')
```

## Shell scriptiong

<a href="../../slides/02a_Unix/03_Scripting_xaringan.html" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/02a_Unix/03_Scripting_xaringan.html#1')
```

## Network

<a href="../../slides/02b_Network/01_Network_SSH_xaringan.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/02b_Network/01_Network_SSH_xaringan.html#1')
```




# References

## Unix general

- A Book for Anyone to Get Started with Unix. [http://seankross.com/the-unix-workbench/](http://seankross.com/the-unix-workbench/), and the GitHub repository, [https://github.com/seankross/the-unix-workbench](https://github.com/seankross/the-unix-workbench)

- Data Coding 101 – Intro To Bash. Four episodes, video. [https://data36.com/data-coding-bash-best-practices/](https://data36.com/data-coding-bash-best-practices/)

- An interactive explainer of any shell command. [http://explainshell.com/](http://explainshell.com/) 

- Unix/Linux command reference sheets. [https://cheat-sheets.s3.amazonaws.com/linux-commands-cheat-sheet-new.pdf](https://cheat-sheets.s3.amazonaws.com/linux-commands-cheat-sheet-new.pdf) and [https://files.fosswire.com/2007/08/fwunixref.pdf](https://files.fosswire.com/2007/08/fwunixref.pdf) 

- Survival guide for Unix newbies. [http://matt.might.net/articles/basic-unix/](http://matt.might.net/articles/basic-unix/) 

- Settling into Unix tutorial. [http://matt.might.net/articles/settling-into-unix/](http://matt.might.net/articles/settling-into-unix/) 

- Shell programming with bash tutorial. [http://matt.might.net/articles/bash-by-example/](http://matt.might.net/articles/bash-by-example/) 

- Master the power of command-line with a list of one-liner gems. [http://www.commandlinefu.com/commands/browse](http://www.commandlinefu.com/commands/browse) 

- "The Unix shell", Software Carpentry. [https://swcarpentry.github.io/shell-novice/](https://swcarpentry.github.io/shell-novice/) 

- A curated list of Terminal frameworks, plugins & resources for command-line interface (CLI) lovers. [http://terminalsare.sexy](http://terminalsare.sexy) and [https://github.com/k4m4/terminals-are-sexy](https://github.com/k4m4/terminals-are-sexy) 

- "Data Science at the Command Line" by Jeroen Janssens, GitHub repository [https://github.com/jeroenjanssens/data-science-at-the-command-line](https://github.com/jeroenjanssens/data-science-at-the-command-line) and an online book [https://www.datascienceatthecommandline.com/](https://www.datascienceatthecommandline.com/)

## Unix in Bioinformatics

- Heng Li's "A Bioinformatician's UNIX Toolbox", [http://lh3lh3.users.sourceforge.net/biounix.shtml](http://lh3lh3.users.sourceforge.net/biounix.shtml)

- Bioinformatics one-liners by Stephen Turner, [https://github.com/stephenturner/oneliners](https://github.com/stephenturner/oneliners)

- Collection of bioinformatics-genomics bash one liners, using awk, sed etc. [https://github.com/crazyhottommy/bioinformatics-one-liners](https://github.com/crazyhottommy/bioinformatics-one-liners)

- Links and references to many genomics and bioinformatics resources, [https://github.com/crazyhottommy/getting-started-with-genomics-tools-and-resources](https://github.com/crazyhottommy/getting-started-with-genomics-tools-and-resources)

## Text manipulation in Unix

- Regular expression, Unix commands, Python quick reference, SQL reference card. [PCfB_Appendices.pdf](misc/PCfB_Appendices.pdf)

- Tutorial to `sed` by Bruce Barnett. [http://www.grymoire.com/Unix/Sed.html](http://www.grymoire.com/Unix/Sed.html) 

- Interactive Vim tutorial. [http://www.openvim.com/](http://www.openvim.com/) 

- Vim reference card. [http://web.mit.edu/merolish/Public/vi-ref.pdf](http://web.mit.edu/merolish/Public/vi-ref.pdf) 