---
title: "Bioconductor"
# linktitle: "1. Introduction to RStudio and Images"
date: "2026-03-16"
class_date: "2026-03-16"
draft: false
output:
  blogdown::html_page:
    toc: false
menu:
  class:
    parent: Class sessions
    weight: 5
type: docs
weight: 5
# pdf: 
# thumb: 
editor_options: 
  chunk_output_type: console
---

## Bioconductor ecosystem for genomic data

<a href="../../slides/05_Bioconductor/01a_Bioconductor_xaringan.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/05_Bioconductor/01a_Bioconductor_xaringan.html#1')
```

**Code:** Biostrings [biostrings.Rmd](../../slides/05_Bioconductor/lab/biostrings.Rmd), [biostrings.html](../../slides/05_Bioconductor/lab/biostrings.html)

<a href="../../slides/05_Bioconductor/01b_Bioconductor_xaringan.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/05_Bioconductor/01b_Bioconductor_xaringan.html#1')
```

Handling genomic data using Bioconductor [PDF](../../slides/05_Bioconductor/genomicRanges.pdf)

**Code:** Genomic Ranges [genomicRanges.Rmd](../../slides/05_Bioconductor/lab/genomicRanges.Rmd)

<a href="../../slides/05_Bioconductor/01c_Bioconductor_xaringan.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/05_Bioconductor/01c_Bioconductor_xaringan.html#1')
```

<a href="../../slides/05_Bioconductor/01d_Bioconductor_xaringan.html#1" target="_blank">Lecture slides in HTML format</a>

```{r echo=FALSE}
knitr::include_url('../../slides/05_Bioconductor/01d_Bioconductor_xaringan.html#1')
```

